Using Path-Dependent Types to Build Type-Safe JavaScript Foreign Function Interfaces

### 2014-01-13 papers/icwe14/icwe14.tex
up

\begin{abstract}
Encoding dynamically typed APIs in statically typed languages can be challenging. We present new ways to encode them in order to preserve type safety and to keep the same expression power.
\end{abstract}

@@

However we observe that, despite these languages are statically typed, their integration of the browser API either is not type safe or gives less expression power to developers. Indeed, integrating an API designed for a dynamically typed language into a statically typed language can be challenging. For instance, the \jscode{createElement} function return type depends on the value of the \jscode{String} parameter it is passed.

This paper reviews some common functions of the browser API, identifies those which are not well encoded in statically typed programming languages and shows new ways to encode them with more type safety while keeping the native API expression power. We find that type parameters are sufficient to achieve this goal and that \emph{path-dependent types} provide an even more convenient encoding for the end developers.

@@

However, in the case of \jscode{getElementsByTagName(name)} and \jscode{createElement(name)}, there is exactly one possible return type for each value of the \jscode{name} parameter: \emph{e.g.} \jscode{getElementsByTagName('input')} always returns a list of \jscode{InputElement} and \jscode{createElement('div')} always returns a \jscode{DivElement}. This characteristic makes it possible to encode each of these functions by defining as many parameterless functions as there are possible tag names, where each function fixes the initial \jscode{name} parameter to be one of the possible values and exposes the corresponding specialized return type. Listing~\ref{lst-create-element-overload} illustrates such an encoding in the Java language.

@@

Table~\ref{table-existing-encodings} summarizes, for each statically typed language, which approach is used to encode each function of the browser API.

@@

We distinguished three approaches to integrate the challenging parts of the browser API into statically typed languages. This section compares these approaches in terms of type safety, expression power and convenience to use from the end developer point of view.

@@

The second approach, consisting in defining as many functions as there are possible return types of a function, is completely type safe but can lead to combinatorial explosion in some situations. Consider for instance listing~\ref{lst-js-comb} defining a JavaScript function that both creates an element and registers an event listener when a given event occurs on it. Note that the event listener is passed both the event and the element. Encoding such a function in a statically typed language using this approach would require to create one function for each combination of tag name and event name.

\begin{figure}
\begin{lstlisting}[label=lst-js-comb,language=JavaScript]
var createAndListenTo = function (tagName, eventName, listener) {
  var element = document.createElement(tagName);
  element.addEventListener(eventName, function (event) {
    listener(event, element);
  });
};
\end{lstlisting}
\end{figure}

@@

\begin{figure}
\begin{lstlisting}[label=lst-js-react,language=JavaScript]
var observe = function (target, name) {
  return function (listener) {
    target.addEventListener(name, listener);
  }
};
\end{lstlisting}
\end{figure}

In summary, browser API integration by existing statically typed languages compiling to JavaScript is either not type safe or not as modular or expressive as the underlying JavaScript API.

@@

In this section we show how we can encode the functions presented previously, in a type safe and convenient way for developers while keeping the same expression power.

@@

Listing~\ref{lst-generics-dom} shows this approach applied to the \jscode{createElement} function: we created a type \scalacode{ElementName<E>}, the \scalacode{name} parameter is not a \scalacode{String} but a \scalacode{ElementName<E>}, and the function returns an \scalacode{E} instead of an \scalacode{Element}. The \scalacode{ElementName<E>} type encodes the relationship between the name of an element and the type of this element\footnote{The type parameter \scalacode{E} is also called a \emph{phantom type}~\cite{leijen1999domain} because \scalacode{ElementName} values never hold a \scalacode{E} value}. For instance, we created a value \scalacode{Input} of type \scalacode{ElementName<InputElement>}. The last two lines shows how this API is used by end developers: by passing the \scalacode{Input} value as parameter to the function, it fixes the type parameter \scalacode{E} to \scalacode{InputElement} so the returned value has the most possible precise type.

\begin{figure}
\begin{lstlisting}[label=lst-generics-dom,language=java]
// --- API Definition
class ElementName<E> {}

@@

\begin{figure}
\begin{lstlisting}[label=lst-generics-events,language=java]
// --- API Definition
class EventName<E> {}

@@

\begin{figure}
\begin{lstlisting}[label=lst-generics-comb,language=java]
<A, B> void createAndListenTo(
    ElementName<A> tagName,
    EventName<B> eventName,
    Function2<A, B, Void> listener) {
  A element = document.createElement(tagName);
  element.addEventListener(eventName, event -> listener.apply(event, element));
}
\end{lstlisting}
\end{figure}

\begin{figure}
\begin{lstlisting}[label=lst-generics-react,language=java]
<A> Function<Function<A, Void>, Void> observe(EventTarget target, EventName<A> name) {
  return listener -> {
    target.addEventListener(name, listener);
  }
}
\end{lstlisting}
\end{figure}

Our encoding makes it possible to implement exactly the same functions as we are able to implement in plain JavaScript, but with type safety.

However, every function taking an element name or an event name as parameter must have type parameters too, making its type signature harder to read.

@@

\begin{figure}
\begin{lstlisting}[label=lst-dt-dom]
trait ElementName {
  type Element
}
object Div extends ElementName { type Element = DivElement }
object Input extends ElementName { type Element = InputElement }

@@

\begin{figure}
\begin{lstlisting}[label=lst-dt-events]
trait EventName {
  type Event
}
object Click extends EventDef { type Event = MouseEvent }

@@

\begin{figure}
\begin{lstlisting}[label=lst-dt-comb]
def createAndListenTo(tagName: ElementName, eventName: EventName, listener: (eventName.Event, tagName.Element) => Unit) = {
  val element = document.createElement(tagName)
  element.addEventListener(eventName)(event => listener(event, element))
}
\end{lstlisting}
\end{figure}

\begin{figure}
\begin{lstlisting}[label=lst-dt-react]
def observe(target: EventTarget, name: EventName) =
  (listener: (name.Event => Unit)) => target.addEventListener(name)(listener)
\end{lstlisting}
\end{figure}

With this encoding, the functions using event names or element names are not anymore cluttered with type parameters.

@@

We presented two ways to encode dynamically typed browser functions in mainstream statically typed languages like Java and Scala, using \emph{generics} or path-dependent types. Our encoding gives more type safety than existing solutions, while keeping the same expression power and modularity level as the native API.

>>>>

\begin{abstract}
Encoding dynamically typed APIs in statically typed languages can be challenging. Incidentally, existing statically typed languages compiling to JavaScript expose its native API in ways that either are not type safe or give less expression power to developers. We present two new ways to encode the challenging parts of the Web browser API such that type safety and expression power are preserved. Our first encoding relies on type parameters and can be implemented in most mainstream languages but drags phantom types up to the use sites. The second encoding does not suffer from this inconvenience but requires dependent types.
\end{abstract}

@@

However we observe that, despite these languages are statically typed, their integration of the browser API either is not type safe or gives less expression power. Indeed, integrating an API designed for a dynamically typed language into a statically typed language can be challenging. For instance, the \jscode{createElement} function return type depends on the value of its parameter: \jscode{createElement('div')} returns a \jscode{DivElement}, \jscode{createElement('input')} returns an \jscode{InputElement}, \emph{etc}.

Existing languages expose this function by making it return an \jscode{Element}, the least upper bound of the types of all the possible returned values, thus loosing type information and requiring users to explicitly downcast the returned value to its expected more precise type. Another way to expose this function consists in exposing several functions, each one fixing the value of the parameter along with its return type: \jscode{createDivElement}, \jscode{createInputElement}, \emph{etc} would be parameterless functions returning a \jscode{DivElement} and an \jscode{InputElement}, respectively. This encoding forces to hard-code the name of the element to create: it can not anymore be a parameter. The first solution is not type safe and the second solution reduces the expression power.

This paper reviews some common functions of the browser API, identifies the patterns that are difficult to encode in static type systems and shows new ways to encode them such that type safety and expression power are preserved. We find that type parameters are sufficient to achieve this goal and that \emph{path-dependent types} provide an even more convenient encoding for the end developers.

@@

However, in the case of \jscode{getElementsByTagName(name)} and \jscode{createElement(name)}, there is exactly one possible return type for each value of the \jscode{name} parameter: \emph{e.g.} \jscode{getElementsByTagName('input')} always returns a list of \jscode{InputElement} and \jscode{createElement('div')} always returns a \jscode{DivElement}. This characteristic makes it possible to encode each of these functions by defining as many parameterless functions as there are possible tag names, where each function fixes the initial \jscode{name} parameter to be one of the possible values and exposes the corresponding specialized return type. Listing~\ref{lst-create-element-overload} illustrates such an encoding in the Java language. Note that the listing gives only the type signature of the functions visible to the end developers. The implementation of the functions is not relevant for our purpose.

@@

Table~\ref{table-existing-encodings} summarizes, for each studied statically typed language, which approach is used to encode each function of the browser API.

@@

We distinguished three approaches to integrate the challenging parts of the browser API into statically typed languages. This section compares these approaches in terms of type safety and expression power.

@@

The second approach, consisting in defining as many functions as there are possible return types of the encoded function, is type safe but leads to a less general API: each function fixes a parameter value of the encoded function, hence being less general. The limits of this approach are better illustrated when one tries to combine several functions. Consider for instance listing~\ref{lst-js-comb} defining a JavaScript function that both find elements and registers an event listener when a given event occurs on them. Note that the event listener is passed both the event and the element: its type depends on both the tag name and the event name. This function could not be implemented if the general \jscode{getElementsByTagName} and \jscode{addEventListener} functions were not available. The best that could be done would be to create one function for each combination of tag name and event name.

\begin{figure}
\begin{lstlisting}[label=lst-js-comb,language=JavaScript,caption={Combination of use of \jscode{getElementsByTagName} and \jscode{addEventListener}}]
var findAndListenTo = function (tagName, eventName, listener) {
  document.getElementsByTagName(tagName).forEach(function (element) {
    element.addEventListener(eventName, function (event) {
      listener(event, element);
    });
  });
};
\end{lstlisting}
\end{figure}

@@

\begin{figure}
\begin{lstlisting}[label=lst-js-react,language=JavaScript,caption={Partial application of \jscode{addEventListener} parameters}]
var observe = function (target, name) {
  return function (listener) {
    target.addEventListener(name, listener);
  }
};
\end{lstlisting}
\end{figure}

[FIXME En fait les deux dernieres approches sont deux cas particuliers d'une meme approche, plus generale, qui consiste a fixer la valeur d'un parametre. Je regroupe ou pas ?]

In summary, browser API integration by existing statically typed languages compiling to JavaScript is either not type safe or not as expressive as the underlying JavaScript API.

@@

In this section we show how we can encode the functions presented previously, in a type safe way while keeping the same expression power.

@@

Listing~\ref{lst-generics-dom} shows this approach applied to the \jscode{createElement} function which return type depends on its \jscode{name} parameter value: a type \scalacode{ElementName<E>} has been created, the type of the \scalacode{name} parameter has been set to \scalacode{ElementName<E>} instead of \scalacode{String}, and the return type of the function is \scalacode{E} instead of \scalacode{Element}. The \scalacode{ElementName<E>} type encodes the relationship between the name of an element and the type of this element\footnote{The type parameter \scalacode{E} is also called a \emph{phantom type}~\cite{leijen1999domain} because \scalacode{ElementName} values never hold a \scalacode{E} value}. For instance, we created a value \scalacode{Input} of type \scalacode{ElementName<InputElement>}. The last two lines shows how this API is used by end developers: passing the \scalacode{Input} value as parameter to the function fixes the type parameter \scalacode{E} to \scalacode{InputElement} so the returned value has the most possible precise type.

\begin{figure}
\begin{lstlisting}[label=lst-generics-dom,language=java,caption={Encoding of the \jscode{createElement} function using type parameters}]
// --- API Definition
class ElementName<E> {}

@@

\begin{figure}
\begin{lstlisting}[label=lst-generics-events,language=java,caption={Encoding of the \jscode{addEventListener} function using type parameters}]
// --- API Definition
class EventName<E> {}

@@

\begin{figure}
\begin{lstlisting}[label=lst-generics-comb,language=java,caption={Combination of \scalacode{getElementsByTagName} and \scalacode{addEventListener} functions encoded using type parameters}]
<A, B> void findAndListenTo(
    ElementName<A> tagName,
    EventName<B> eventName,
    Function2<A, B, Void> listener) {
  for (A element : document.getElementsByTagName(tagName)) {
    element.addEventListener(eventName, event -> listener.apply(event, element));
  }
}
\end{lstlisting}
\end{figure}

\begin{figure}
\begin{lstlisting}[label=lst-generics-react,language=java,caption={Partial application of \scalacode{addEventListener} encoded with type parameters}]
<A> Function<Function<A, Void>, Void> observe(EventTarget target, EventName<A> name) {
  return listener -> {
    target.addEventListener(name, listener);
  }
}
\end{lstlisting}
\end{figure}

Our encoding is type safe and gives as much expression power as the native API since it makes it possible to implement exactly the same functions as we are able to implement in plain JavaScript.

However, every function taking an element name or an event name as parameter has its type signature cluttered with phantom types (extra type parameters).

@@

\begin{figure}
\begin{lstlisting}[label=lst-dt-dom,caption={Encoding of \jscode{createElement} using path-dependent types}]
trait ElementName {
  type Element
}
object Div extends ElementName { type Element = DivElement }
object Input extends ElementName { type Element = InputElement }

@@

\begin{figure}
\begin{lstlisting}[label=lst-dt-events,caption={Encoding of \jscode{addEventListener} using path-dependent types}]
trait EventName {
  type Event
}
object Click extends EventDef { type Event = MouseEvent }

@@

\begin{figure}
\begin{lstlisting}[label=lst-dt-comb,caption={Combination of \scalacode{getElementsByTagName} and \scalacode{addEventListener} using path-dependent types}]
def findAndListenTo(tagName: ElementName, eventName: EventName, listener: (eventName.Event, tagName.Element) => Unit) = {
  for (element <- document.getElementsByTagName(tagName) {
    element.addEventListener(eventName)(event => listener(event, element))
  }
}
\end{lstlisting}
\end{figure}

\begin{figure}
\begin{lstlisting}[label=lst-dt-react,caption={Partial application of \scalacode{addEventListener} using path-dependent types}]
def observe(target: EventTarget, name: EventName) =
  (listener: (name.Event => Unit)) => target.addEventListener(name)(listener)
\end{lstlisting}
\end{figure}

With this encoding, the functions using event names or element names are not anymore cluttered with phantom types, and type safety is still preserved.

@@

We presented two ways to encode dynamically typed browser functions in mainstream statically typed languages like Java and Scala, using type parameters or path-dependent types. Our encoding gives more type safety than existing solutions, while keeping the same expression power as the native API.

We argue that, if industry shippers want to write their Web applications in statically typed languages, dependent types are going to be the most desired feature of these languages.
